Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-17 MS/ms.Rmd
this looks great - just a few comments

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Meta-analysis' core objective and strength is to distill published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance), and which can be compared across broad sets of research designs, study systems and species. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-regression models to both account for factors driving differences across studies and make them more comparable. In addition, we formalise effect size measures that provide comparative physiologists with a means to directly make effect measures comparable, removing 'nuisance heterogeneity' without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be usefully applied to a variety of different contexts to derive 'new' effects sizes and sampling variances. The effect sizes and their corresponding sampling variances that we describe should help physiologists deal with common treatment differences across studies. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses are part of a systematic review and involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. A meta-analysis should follow and report upon this entire process carefully to ensure the review process and data collation, are as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of a research field there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. A critically important feature of a meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased [@Morrissey2016a].

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, having highly heterogeneous effects "...is the spice of life" [p.g., 519; @Cooperetal2019] because it provides opportunities to explore the reasons for why effects vary within and across studies [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity in a meta-analysis is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. Finally, we describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating the 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. It might already be clear that these factors will have an impact on the magnitude of an effect, but they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

@@

Before embarking on "meta-analytic enlightenment" it is worth first understanding how comparative physiologists are currently conducting meta-analyses / comparative analysis.  As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

#### *Current Meta-analytic Practices in Comparative Physiology*

```{r setup2, include=FALSE}
rm(list=ls())

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`\% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`\% of studies), standardised mean differences (SMD) and log response ratios (lnRR) (`r with(effects, effects[effect == "SMD","prop"])*100`\% and `r with(effects, effects[effect == "lnRR","prop"])*100`\%, respectively; Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`\% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`\% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B); usually through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`\% of studies used a weighted meta-analytic model, however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`\% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not actually weight their meta-analytic mean. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`\% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortaunte given how important these measures are to interpreting effects.

@@

eqns_yi <- c("M",
             "\\ln{SD} + \\frac{1}{2(N - 1)}",
             # Senior will have a better version but this is good for now
             "\\ln \\left( \\frac{M_{1}}{M_{2}}\\right)",
             "\\frac{\\left( M_{2} - M_{1}\\right)}{SD_{p}}J",
             # we need to spell out SD_p somewhere
             "\\frac{1}{2}log\\left(\\frac{1+r}{1-r}\\right)")
             #"log\\left(\\frac{y_{i2}+0.5}{n_{i2}-y_{i2}+0.5}\\right) ",
             #"- log\\left(\\frac{y_{i1}+0.5}{n_{i1}-y_{i1}+0.5}\\right)")

@@

As can be seen in Table \@ref(tab:tablea) none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to control for various sources of 'nuisance heterogeneity'. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common, well understood effect measure of interest to comparative physiologists, $Q_{10}$, how we can formalise a traditional effect size, such as $lnRR$, to capture this 'nuisance variation'.

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A common experiment in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using the effect size measures in Table \@ref(tab:tablea) would result in the effect sizes varying with the temperatures applied within and across across studies contributing to 'comparability' issues. For example, an effect size for a temperature manipulation of 10°C would be larger than an effect size for a temperature manipulation of 5°C (see Figure \@ref(fig:fig1)). As such, it is common to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

Our example using $Q_{10}$, a well known effect measure in comparative physiology, shows how common effect size's can be re-formalised to account for 'nuisance heterogeneity', in this case temperature. Of course, in doing this we are making assumptions about the nature of temperature effects on an effect size. Nonetheless, we can apply similar approaches to other commonly used effect size metrics. We describe more generally how this can be done in Box 1, applying similar principles to standardise temperature differences for slopes and for standardised mean differences.

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of animals (or plants) (e.g., may be because many traits are measured on the same sets of animals) then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying \textbf{I} to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

@@

Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of 'nuisance' variation, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (https://github.com/daniel1noble/metaAidR).

>>>>

Meta-analysis is a tool that provides comparative physiologists with powerful and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. <!--- I took "unbiased" out as this is not quite true - it can be till biased ----> Meta-analysis' core objective and strength is to distill published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance), and which can be compared across broad sets of research designs, study systems and species. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-regression models to both account for factors driving differences across studies and make them more comparable. In addition, we formalise effect size measures that provide comparative physiologists with a means to directly make effect measures comparable, removing 'nuisance heterogeneity' without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be usefully applied to a variety of different contexts to derive 'new' effects sizes and sampling variances. The effect sizes and their corresponding sampling variances that we describe should help physiologists deal with common treatment differences across studies. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses are part of a systematic review and involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. A meta-analysis should follow and report upon this entire process carefully to ensure the review process and data collation, are as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of a research field there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. A critically important feature of a meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies in the analysis and make overall estimates more precise (note that this statement is only completely true for the fixed-effect meta-analysis and that actual weighting is more complex in the model introduced below although this is a good intuition for all of us to have). <!--- just being careful ----> While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself may be unbiased [@Morrissey2016a].<!--- as said, unweighted will be more suseptible to publication bias ---->

Despite its widespread adoption and well-established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, having highly heterogeneous effects "...is the spice of life" [p.g., 519; @Cooperetal2019] because it provides opportunities to explore the reasons for why effects vary within and across studies [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity in a meta-analysis is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. <!--- I guss if we map the road like this - it may be fine as readers have been warned ---> Then, we present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We also formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. Finally, we describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

<!--- Dan - this is really good below --->
Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating the 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. It might already be clear that these factors will have an impact on the magnitude of an effect, but they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

@@

Before embarking on "meta-analytic enlightenment", it is worth first understanding how comparative physiologists are currently conducting meta-analyses / comparative analysis. As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

#### *Current Meta-analytic Practices in Comparative Physiology*
<!--- the fig 2 the label "Others" - should they come at the bottom??? Also, for the method - you may want to add all these data, survey etc can be found at the github repo??? --->
```{r setup2, include=FALSE}
rm(list=ls())

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (SMD) and log response ratios (lnRR) (`r with(effects, effects[effect == "SMD","prop"])*100`% and `r with(effects, effects[effect == "lnRR","prop"])*100`%, respectively; Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B); usually through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model, however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not actually weight their meta-analytic mean. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortaunte given how important these measures are to interpreting effects.

@@

eqns_yi <- c("M",
             "\\ln{SD} + \\frac{1}{2(N - 1)}",
             # Senior will have a better version but this is good for now
             "\\ln \\left( \\frac{M_{1}}{M_{2}}\\right)",
             "\\frac{\\left( M_{2} - M_{1}\\right)}{SD_{p}}J",
             # we need to spell out SD_p somewhere
             "\\frac{1}{2}log\\left( \\frac{1+r}{1-r}\\ right)")
             #"log\\left(\\frac{y_{i2}+0.5}{n_{i2}-y_{i2}+0.5}\\right) ",
             #"- log\\left(\\frac{y_{i1}+0.5}{n_{i1}-y_{i1}+0.5}\\right)")

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to control for various sources of 'nuisance heterogeneity'. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common, well understood effect measure of interest to comparative physiologists, $Q_{10}$, how we can formalise a traditional effect size, such as $lnRR$, to capture this 'nuisance variation'.

#### *Comparing Changes in Mean Physiological Rates,* $Q_{10}$

A common experiment in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using the effect size measures in Table \@ref(tab:tablea) would result in the effect sizes varying with the temperatures applied within and across across studies contributing to 'comparability' issues. For example, an effect size for a temperature manipulation of 10°C would be larger than an effect size for a temperature manipulation of 5°C (see Figure \@ref(fig:fig1)). As such, it is common to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not seem to have associated sampling variances derived for $Q_{10}$ in the literature, making it challenging to make use of the full power of weighted meta-analytic models. <!--- I am bing a bit more careful here ---> Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

Our example using $Q_{10}$, a well-known effect measure in comparative physiology, shows how common effect size's can be re-formalised to account for 'nuisance heterogeneity', in this case temperature. Of course, in doing this we are making assumptions about the nature of temperature effects on an effect size. Nonetheless, we can apply similar approaches to other commonly used effect size metrics. We describe more generally how this can be done in Box 1, applying similar principles to standardise temperature differences for slopes and for standardised mean differences.

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of animals (or plants) (e.g., may be because many traits are measured on the same sets of animals) then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

@@

Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010; @Nakagawa2017]. If we want to formally assess how much effect size variation is the result of 'nuisance' variation, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).
